---
title: "P8451 Machine Learning in Public Health - Assignment 1"
output: html_document
date: "2023-1-24"
---

In preparation for the problems below, we will load the following libraries:

```{r load_libraries}
library(tidyverse)
library(dplyr)
library(psych)
```

# Question 1

## 1.1 Data Import and Cleaning

We will begin by importing the dataset used by authors Patricio et al for the development of a biomarker-based prediction model to identify women with breast cancer. To do so, we will first read in the data by using the `read_csv` function, then apply the `clean_names` function to quickly clean the data. 

```{r}
breast_cancer_data = 
  read_csv("./bcdata_Assignment1.csv") %>% 
  janitor::clean_names()
```

## 1.2 Tabulating Summaries of Quantitative Features

To generate a table providing the summaries of the quantitative features of the dataset, we can apply the `describe` function, which generates a variety of summary statistics. From this output, we will select the following: mean, median, minimum value, and maximum value. 

```{r}
describe(breast_cancer_data) %>% 
  select("mean", "median", "min", "max") %>% 
  knitr::kable(col.names = c('Mean', 'Median', 'Minimum Value', 'Maximum Value'))
```

# Question 2 

We will now recode the existing BMI variable into the following WHO-defined categories:

 - Severely Underweight - BMI less than 16.5 kg/m^2
 - Underweight - BMI under 18.5 kg/m^2
 - Normal Weight - BMI greater than or equal to 18.5 to 24.9 kg/m^2
 - Overweight – BMI greater than or equal to 25 to 29.9 kg/m^2
 - Obesity Class I – BMI 30 to 34.9 kg/m^2
 - Obesity Class II – BMI 35 to 39.9 kg/m^2
 - Obesity Class III – BMI greater than or equal to 40 kg/m^2
 
This process involves converting the existing `bmi` variable from a numeric to factor variable with 7 levels, using the `as.factor` function, as well as a series of `ifelse` statements. 
 
```{r}
breast_cancer_data$bmi_category <- as.factor(ifelse(breast_cancer_data$bmi < 16.5, 'Severely Underweight',
                                             ifelse(breast_cancer_data$bmi < 18.5, 'Underweight', 
                                             ifelse(breast_cancer_data$bmi < 25, 'Normal Weight',
                                             ifelse(breast_cancer_data$bmi < 30, 'Overweight',
                                             ifelse(breast_cancer_data$bmi < 35, 'Obesity Class I',
                                             ifelse(breast_cancer_data$bmi < 40, 'Obesity Class II',
                                             'Obesity Class III')))))))
```

# Question 3

Using the dataset generated from Question 2, we can generate a bar chart depicting the proportion of breast cancer cases and controls within each BMI category. 

```{r}
breast_cancer_data %>% 
  group_by(bmi_category) %>%
  mutate(n = n(),
         cases = sum(classification - 1),
         prop_cases = cases / n) %>% 
  ggplot(aes(x = reorder(bmi_category, prop_cases), fill = factor(classification))) +
  geom_bar(position = "fill") +
  labs(
    title = "Proportions of Breast Cancer Cases & Controls within \nWHO-Defined BMI Categories", 
    x = "BMI Categories", 
    y = "Proportion", 
    fill = "Classification") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom") +
  scale_fill_discrete(labels = c('Control', 'Case'))
```

# Question 4

Below we construct a logistic regression model using breast cancer classification as the outcome of interest, and glucose, HOMA-IR, leptin, continuous BMI, and age as the independent variables of interest.

```{r}
breast_cancer_data = breast_cancer_data %>% 
  mutate(classification = as.factor(recode(classification, 
                                           "1" = 0,
                                           "2" = 1)))

classification_logistic = 
  breast_cancer_data %>% 
  glm(classification ~ glucose + homa + leptin + bmi + age, data = ., family = binomial())


classification_logistic %>% 
  broom::tidy() %>% 
  mutate(lower_CI = estimate - 1.96*std.error, 
         upper_CI = estimate + 1.96*std.error) %>% 
  select(term, estimate, lower_CI, upper_CI, p.value) %>% 
  knitr::kable(col.names = c('Term', 'beta Estimate', 'Lower 95% CI', 'Upper 95% CI', "p-value"), digits = 3)
```

The beta estimate associated with a 1-unit change in _HOMA-IR_ is __0.274__. The 95% confidence interval surrounding this beta estimate is __(-0.063, 0.611)__. 

# Question 5

Below we construct linear regression model using insulin as the outcome of interest, and continuous BMI, age, and glucose as the independent variables of interest.

```{r}
insulin_linear = 
  breast_cancer_data %>% 
  lm(insulin ~ bmi + age + glucose, data = .)

insulin_linear %>% 
  broom::tidy() %>% 
  mutate(lower_CI = estimate - 1.96*std.error, 
         upper_CI = estimate + 1.96*std.error) %>% 
  select(term, estimate, lower_CI, upper_CI, p.value) %>% 
  knitr::kable(col.names = c('Term', 'beta Estimate', 'Lower 95% CI', 'Upper 95% CI', "p-value"), digits = 3)
```

The beta estimate associated with a 1-unit change in _age_ is __-0.054__. The 95% confidence interval surrounding this beta estimate is __(-0.156, 0.048)__. 
